A Kochen-Specker system has at least 21 vertices\\ {\small (extended abstract)}

### 2014-04-14 main.tex
bla

Quantum Mechanics only predicts the chance, given the initial configuration
of the field, whether the particle will deviate.
Its probabilistic prediction has been thoroughly tested.
One wonders: is there a deterministic theory predicting the
outcome of this experiment?

Kochen and Specker proved that such a theory cannot satisfy:
% TODO note contextuality
\begin{spin}
    Given three pairwise orthogonal directions.
    In exactly one of the directions, the particle will not deviate.
\end{spin}
Their argument is based on the existence of a Kochen-Specker system.
\begin{dfn}
    A \keyword{Kochen-Specker (KS) system} is
    a finite set of points on the sphere
    for which each pair is not antipodal and
    there is no~\keyword{010-coloring}.
    A $010$-coloring is a~$\{0,1\}$-coloring of the points such that
        \footnote{
                In other papers, like \cite{aow11},
                the~$0$ and~$1$ are swapped:
                101-colorings.
                Then the colors correspond directly to the squared measurement
                outcome of the spin.}
    \begin{enumerate}
        \item
            no pair of orthogonal points are both colored~$1$ and
        \item
            of three pairwise orthogonal points exactly one is colored~$1$;
            or alternatively: they are colored~$0$, $1$ and~$0$ in some order.
    \end{enumerate}
\end{dfn}
% TODO shorten this proof.
A point on the sphere obviously corresponds to a direction in space.
Because of this, the term point, vector and direction
can be used interchangeably. Antipodal points correspond to oppositve
vectors and these span the same direction in space.

Suppose there is a KS system and  a deterministic theory satisfying
the SPIN Axiom.
Then we color a point of the system~$0$,
whenever the theory predicts that the particle will deviate
if the magnetic field is directed along that point and~$1$ otherwise.
Given two orthogonal points of the system,
we can find a third point orthogonal to both of them.
The SPIN axiom implies exactly one of them is colored~$1$.
Thus they cannot both be colored~$1$.
And similarly, given three pairwise orthogonal vectors in the system,
the SPIN axiom implies exactly one of them is colored~$1$.
Hence the KS system is 010-colorable, quod non.  Thus the deterministic theory
cannot satisfy the SPIN Axiom.

@@

First, in Subsection~\ref{sec:ksgraphs},
we repeat a part of AOW's work, in particular the reduction of
KS systems to graphs.
The bottleneck of their search was the sheer number of graphs
and the deciding embeddability.

@@

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasable as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.
% TODO OEIS A000088
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph does not contain a square.
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthognal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$ vertices.
We can restrict ourselves to connected graphs.
% TODO OEIS A006786
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
We have verified the main result of \cite{aow11}:
\begin{comp}
The unique non-010-colorable squarefree connected graph on~$17$
vertices is. TODO
It is not embeddable, see TODO, and hence a Kochen-Specker
system has at least 18 points.
\end{comp}

@@

Now consider a leaf~$A$ in the biconnected component graph~$B(G)$.
Let~$B$ be the union of the remaining biconnected components.
There is exactly one pair of vertices~$(a,b)$
such that~$a \in A$, $b \in B$ and $a$ is adjacent to~$b$.

@@

These two rotations yield a new set of points on the sphere,
that demonstrate~$G'$ is embeddable. 
\end{proof}
We believe a minimal KS graph is also triconnected.
Although these arguments are pleasing,
they are of little use:
% TODO Add argument or computation for {2,3}-connected
% TODO note that restricting to (bi/tri)-connected is a waste of time
% TODO add graph 
\begin{comp}
    There are five non-isomorphic minimal
    squarefree connected graphs
    with minimal vertex order 3 and they have 10 vertices.
\end{comp}
\begin{cor}
    Any unconnected
    squarefree graph with minimal vertex order 3
    has at least 20 vertices, for it has two connected components,
    each with at least 10 vertices.
\end{cor}
% TODO is core x years a proper expression?
Now, we can state our main computation, which took roughly a week
    on a 64-core Opteron 6276.
\begin{comp}
    Let~$C_n$ denote the number of non-010 colorable squarefree
    graphs with minimal vertex order 3 on~$n$ nodes.  Then:

@@

In~\cite{aow11}, Arends, Wampler and Ouaknine discuss several
computer-aided methods
to test embeddability of a graph.  None of the methods could decide
of all graphs considered, whether they were embeddable or not.
% TODO give a small description of their deficiencies
% TODO give intermediate computation.

We propose a new method,
which of all graphs we considered,
could decide
whether they were embeddable or not.
First we give a pen-and-paper example.
\begin{prop}\label{prop:unemb-10-2}
The graph in Figure~\ref{fig:unemb-10-2}
is not embeddable.
\end{prop}
\begin{proof}
Suppose it is embeddable.
Consider~$p_1$.
It is orthogonal to both~$a$ and~$v$.
$a$ and~$v$ are not collinear,
hence~$p_1$ must be collinear to~$v \times a$,
the cross-product of~$v$ and~$a$.
Similarly, $p_2$ is collinear to~$v \times p_1 = v \times (v \times a)$.
Continuing in this fashion,
we see that
\begin{equation}\label{eq:ue1}
    a \text{ is collinear to }
    x \times (x \times( w \times (w\times (v \times (v \times a))))).
\end{equation}
Now, we may assume that~$z=(0,0,1)$ and $x=(1,0,0)$.
Thus: $v=(v_1,v_2,0)$;
$w = (w_1,w_2,0)$
and $a = (0, a_2,a_3)$ for some~$-1 \leq v_1,v_2,w_1,w_2,a_2,a_3 \leq 1$
with~$v_1^2+v_2^2 = 1$; $w_1^2+w_2^2=1$ and~$a_2^2 + a_3^2=1$.
Now, \eqref{eq:ue1} becomes:
\begin{equation*}
\begin{pmatrix}
0\\
a_2\\
a_3\\
\end{pmatrix}
\text{ is collinear to }
\begin{pmatrix}
0\\
-a_2 v_1w_2 (v_1w_1 + v_2w_2) \\
-a_3 (v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2)\\
\end{pmatrix}.
\end{equation*}
And thus
\begin{align*}
v_1w_2 \left<v,w\right> & =
v_1w_2 (v_1w_1 + v_2w_2) \\
& = v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2 \\
& = (v_1^2 + v_2^2)w_1^2 + (v_1^2 + v_2^2)w_2^2 \\
& = w_1^2 + w_2^2 = 1.
\end{align*}
Since~$v$ and~$w$ are not collinear,
we have by Cauchy-Schwarz $|\left<v,w\right>| < 1$.
Recall~$|v_1|, |w_2| \leq 1$.
Thus: $|v_1w_2\left<v,w\right>| < 1$.
Contradiction.
\end{proof}

@@

\begin{algorithm}
    \While{there are unassigned vertices}{
        pick an unassigned vertex~$v$\;
        set~$V(v)=v$\;
        mark~$v$ as free\;
        \While{there are unassigned vertices adjacent to two
                different assigned vertices}{
            pick such a vertex~$w$ adjacent to the assigned ~$w_1$ and~$w_2$\;
            set~$V(w)=V(w_1) \times V(w_2)$\;
            mark edges~$(v,w_1)$ and~$(v,w_2)$ as accounted for\;
        }   
    }
    \For{each pair of vertices~$(v_1, v_2)$}{
        \If{$(v_1,v_2)$ is not an edge} {
            record requirement:~``$V(v_1)$ is not collinear to $V(v_2)$''\;
        }
    }
    \For{each edge~$(v_1,v_2)$ not accounted for}{
        record requirement:~``$V(v_1)$ is orthogonal to~$V(v_2)$''\;
    }
\end{algorithm}
At two points in the algorithm, there is a choice which vertex to pick.
Depending on the vertices chosen, the number of recorded requirements
and free points may significantly.  By considering all possible choices,
one can find the one with least free points.

>>>>

Quantum Mechanics only predicts the probability, given the configuration
of the field, whether the particle will deviate.
Its probabilistic prediction has been thoroughly tested.
One wonders: is there a deterministic noncontextual theory predicting the
outcome of this experiment?

Kochen and Specker proved that such a theory cannot satisfy:
% TODO note contextuality
\begin{spin}
    Given three pairwise orthogonal directions.
    In exactly one of the directions, the particle will not deviate.
\end{spin}
Their argument is based on the existence of a Kochen-Specker system.
\begin{dfn}
    A \keyword{Kochen-Specker (KS) system} is
    a finite set of points on the sphere
    for which each pair is not antipodal and
    there is no~\keyword{010-coloring}.
    A $010$-coloring is a~$\{0,1\}$-coloring of the points such that
        \footnote{
                In other papers, like \cite{aow11},
                the~$0$ and~$1$ are swapped, they consider 101-colorings.
                These colorings are of course equivalent and the difference
                arises from considering either squared spin measurements $S^2$, or $1-S^2$.
                %Then the colors correspond directly to the squared measurement
               	%outcome of the spin.
               	}
    \begin{enumerate}
        \item
            no pair of orthogonal points are both colored~$1$ and
        \item
            of three pairwise orthogonal points exactly one is colored~$1$;
            or alternatively: they are colored~$0$, $1$ and~$0$ in some order.
    \end{enumerate}
\end{dfn}
% TODO shorten this proof.
A point on the sphere obviously corresponds to a direction in space.
Because of this, the term point, vector and direction
can be used interchangeably. Antipodal points correspond to opposite
vectors and these span the same direction in space.

Suppose there is a KS system and a noncontextual deterministic theory satisfying
the SPIN Axiom.
Then we color a point of this system~$0$,
whenever this theory predicts that the particle will deviate
if the spin is measured in the direction corresponding to that point, and~$1$ otherwise.
Given two orthogonal points of the system,
we can find a third point orthogonal to both of them.
The SPIN axiom implies exactly one of them is colored~$1$, so they cannot both be colored~$1$.
Similarly, given three pairwise orthogonal vectors in the system,
the SPIN axiom implies exactly one of them is colored~$1$.
Hence there would be a 010-coloring of the KS system, quod non. Therefore a deterministic noncontextual theory cannot satisfy the SPIN Axiom.\\

@@

First, in Subsection~\ref{sec:ksgraphs},
we repeat a part of AOW's work, in particular the reduction of
KS systems to graphs.
The bottleneck of their search was the sheer number of graphs
and the deciding whether such graphs are embeddable.

@@

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasible as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.
% TODO OEIS A000088
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph does not contain a square.
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthognal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$ vertices.
We can restrict ourselves to connected graphs.
% TODO OEIS A006786
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
We have verified the main result of \cite{aow11}:
\begin{comp}
There is a unique non-010-colorable squarefree connected graph on~$17$ or less
vertices. It looks as follows TODO
It is not embeddable, see TODO, and hence a Kochen-Specker
system has at least 18 points.
\end{comp}

@@

Now consider a leaf~$A$ in the biconnected component graph~$B(G)$. That is, $A$ is a component which is only connected to one other component.
Let~$B$ be the union of the remaining biconnected components.
There is exactly one pair of vertices~$(a,b)$
such that~$a \in A$, $b \in B$ and $a$ is adjacent to~$b$.

@@

These two rotations yield a new set of points on the sphere,
that demonstrate~$G'$ is embeddable.
Therefore, in generating all relevant graphs, we find both $G$ and $G'$. Since they have the same order, we might as well ignore the graph $G$. This shows it is sufficient to consider only biconnected graphs.
\end{proof}
We believe a minimal KS graph is also triconnected.
Although these arguments are pleasing,
they are of little use:
% TODO Add argument or computation for {2,3}-connected
% TODO note that restricting to (bi/tri)-connected is a waste of time
% TODO add graph 
\begin{comp}
    There are five non-isomorphic minimal
    squarefree connected graphs
    with minimal vertex order 3 and they have 10 vertices.
\end{comp}
\begin{cor}
    Any unconnected
    squarefree graph with minimal vertex order 3
    has at least 20 vertices, for it has two connected components,
    each with at least 10 vertices.
\end{cor}

% TODO is core x years a proper expression?
Now, we can state our main computation, which took roughly a week
    on a 64-core Opteron 6276.
\begin{comp}
    Let~$C_n$ denote the number of non-010 colorable squarefree
    graphs with minimal vertex order 3 on~$n$ nodes.  Then:

@@

In~\cite{aow11}, Arends, Wampler and Ouaknine discuss several
computer-aided methods
to test embeddability of a graph.  None of these methods could decide
for all graphs considered, whether they were embeddable or not.
% TODO give a small description of their deficiencies
% TODO give intermediate computation.

We propose a new method,
which for all graphs we considered,
could decide
whether they were embeddable or not.
First we give a pen-and-paper example.
\begin{prop}\label{prop:unemb-10-2}
The graph in Figure~\ref{fig:unemb-10-2}
is not embeddable.
\end{prop}
\begin{proof}
Suppose it is embeddable.
Consider~$p_1$.
It is orthogonal to both~$a$ and~$v$.
$a$ and~$v$ are not collinear,
hence~$p_1$ must be collinear to~$v \times a$,
the cross-product of~$v$ and~$a$.
Similarly, $p_2$ is collinear to~$v \times p_1 = v \times (v \times a)$.
Continuing in this fashion,
we see that
\begin{equation}\label{eq:ue1}
    a \text{ is collinear to }
    x \times (x \times( w \times (w\times (v \times (v \times a))))).
\end{equation}
Now, we may assume that~$z=(0,0,1)$ and $x=(1,0,0)$.
Thus: $v=(v_1,v_2,0)$;
$w = (w_1,w_2,0)$
and $a = (0, a_2,a_3)$ for some~$-1 < v_1,v_2,w_1,w_2,a_2,a_3 < 1$,
with~$v_1^2+v_2^2 = 1$; $w_1^2+w_2^2=1$ and~$a_2^2 + a_3^2=1$.
Now, \eqref{eq:ue1} becomes:
\begin{equation*}
\begin{pmatrix}
0\\
a_2\\
a_3\\
\end{pmatrix}
\text{ is collinear to }
\begin{pmatrix}
0\\
-a_2 v_1w_2 (v_1w_1 + v_2w_2) \\
-a_3 (v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2)\\
\end{pmatrix}.
\end{equation*}
And therefore
\begin{align*}
v_1w_2 \left<v,w\right> & =
v_1w_2 (v_1w_1 + v_2w_2) \\
& = v_1^2 w_1^2 + v_1^2 w_2^2 + v_2^2w_1^2 + v_2^2w_2^2 \\
& = (v_1^2 + v_2^2)w_1^2 + (v_1^2 + v_2^2)w_2^2 \\
& = w_1^2 + w_2^2 = 1.
\end{align*}
Since~$v$ and~$w$ are not collinear,
we have by Cauchy-Schwarz $|\left<v,w\right>| < 1$.
Recall~$|v_1|, |w_2| \leq 1$.
Thus: $|v_1w_2\left<v,w\right>| < 1$.
Contradiction.
\end{proof}

@@

\begin{algorithm}
    \While{there are unassigned vertices}{
        pick an unassigned vertex~$v$\;
        set~$V(v)=v$\;
        mark~$v$ as free\;
        \While{there are unassigned vertices adjacent to two
                different assigned vertices}{
            pick such a vertex~$w$ adjacent to the assigned ~$w_1$ and~$w_2$\;
            set~$V(w)=V(w_1) \times V(w_2)$\;
            mark edges~$(v,w_1)$ and~$(v,w_2)$ as accounted for\;
        }   
    }
    \For{each pair of vertices~$(v_1, v_2)$}{
        \If{$(v_1,v_2)$ is not an edge} {
            record requirement:~``$V(v_1)$ is not collinear to $V(v_2)$''\;
        }
    }
    \For{each edge~$(v_1,v_2)$ not accounted for}{
        record requirement:~``$V(v_1)$ is orthogonal to~$V(v_2)$''\;
    }
\end{algorithm}
At two points in the algorithm, there is a choice which vertex to pick.
Depending on the vertices chosen, the number of recorded requirements
and free points may significantly vary. By considering all possible choices,
one can find the one with least free points.
